---
title: "Bayesian Hierarchical Modeling of NBA Shot-Making Probability"
author: "Rex Nelson, William Ken Li, Andrew Cho"
format:
  html:
    toc: true
    toc-depth: 3
    theme: darkly
    code-fold: true
execute:
  warning: false
  message: false
  echo: true
  code-fold: true
---

# Research question

Across all levels of basketball, shot attempts are not equally difficult for every player or every situation.  For example, layups, mid-range shots, and threes all have different expected outcomes.  Also, players differ in shooting ability, and players with small number of attempts are not effectively evaluated from raw field goal percentage alone.

Our research question is:

> How does shot context and player-specific shooting ability affect the probability that an NBA player makes a field goal attempt?

To address this question, we are going to analyze shot-level NBA data using Bayesian hierarchical logistic regression model.  The response variable is binary whether the field goal was made or missed.  We will use shot distance, shot type, shot zone, period, time remaining in the period, score margin, home/away status, and player profile to predict the probability of the field goal being made.  Using the hierarchical model, we hallow each player to have their own shooting ability parameter while shrinking players towards league average.  This is very important in our context since it allows players with fewer shot attempts to be treated on the same level as players with high volume which fixes the unreliability issue with raw field goal percentage.

# Statistical background

For each shot attempt (i = 1, \ldots, n), we define

$$Y_i =
\begin{cases}
1, & \text{if shot } i \text{ was made} \\
0, & \text{if shot } i \text{ was missed.}
\end{cases}$$

Now we model

$$Y_i \mid p_i \sim \text{Bernoulli}(p_i),$$

where (p_i) is the probability that the i'th shot is made.

Because (p_i) must lie between 0 and 1, we chose a logistic regression model:

$$\text{logit}(p_i) = \log\left(\frac{p_i}{1-p_i}\right) = \eta_i.$$

A non-hierarchical model would use

$$\eta_i =
\alpha
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 \, I(\text{3PT}_i)
+
\beta_3 \, I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{zone}_i
+
\beta_6 \,\text{clock}_{z,i}
+
\beta_7 \,\text{margin}_{z,i}.$$

However, this ignores player-to-player variation. We instead use a hierarchical model:

$$\eta_i =
\alpha
+
u_{\text{player}[i]}
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 \, I(\text{3PT}_i)
+
\beta_3 \, I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{zone}_i
+
\beta_6 \,\text{clock}_{z,i}
+
\beta_7 \,\text{margin}_{z,i}.$$

And model the player effects as

$$u_j \sim \text{Normal}(0, \sigma_{\text{player}}),$$

where (u_j) is the adjusted shooting ability of player (j), and (\sigma\_{\text{player}}) describes how much true shooting ability varies across players after controlling for shot context.

Our Bayesian model specifies priors for all unknown parameters and we will use posterior inference:

$$p(\theta \mid y) \propto p(y \mid \theta)p(\theta),$$

where (p(y \mid \theta)) is the likelihood and (p(\theta)) is the prior distribution.

# Bayesian Methodologies

This project goes beyond a standard logistic regression because it leverages a hierarchical Bayesian model. The Bayesian framework is especially useful here for three reasons:

First, the model can estimate uncertainty using posterior distributions not just through joint estiamtes and standard errors.

Second, using partial pooling, we can stabilize player estimates meaning that for a player who makes 6 out of 10 shots we should not necessarily say that they are a better shooter than a player who makes 420 out of 900 shots.  The hierarchical prior shrinks low sample players towards league average.

Third, Bayesian posterior prediction increases interpretability and allows us to answer questions like:

- How much does increasing shot distance reduce the probability of a make?
- What is the predicted probability that a certain player makes a three?
- Which players are overperforming their expected make expectation?

# Import Packages

```{r}
library(tidyverse)
library(hoopR)
library(brms)
library(bayesplot)
library(posterior)
library(lme4)
library(broom.mixed)
library(loo)
library(gt)
library(janitor)
library(forcats)
library(purrr)
```

```{r}
theme_set(theme_minimal(base_size = 13))
options(mc.cores = parallel::detectCores())
set.seed(415)
```

# Data source

We are using play-by-play data acquired through the hoopR package.  We will use load_nba_pbp() and load_nba_player_box() to obtain the play-by-play data and recover player display names from player ID's when available.

Our project proposal leverages play-by-play data which is better suited for a hierarchical model since it provides us with consistent shot-level coordinates on the floor and game context.

# Data collection

The following code pulls the play-by-play data and filters it down to field goal attempts.  Then, we performed a join since the athlete_display_name columns were not complete.  The data also does not contain a shot_distance column so we created an approximate shot distance formula from the play text where posible and used the integrated shot coordinates as a fall-back.

```{r}
season_used <- 2024

pbp_raw <- load_nba_pbp(seasons = season_used) |>
  clean_names()

player_lookup <- tryCatch(
  load_nba_player_box(seasons = season_used) |>
    clean_names() |>
    distinct(athlete_id, athlete_display_name),
  error = function(e) {
    tibble(
      athlete_id = integer(),
      athlete_display_name = character()
    )
  }
)

shots_raw <- pbp_raw |>
  mutate(
    shot_text = str_to_lower(coalesce(text, "")),
    parsed_distance = str_extract(shot_text, "\\d+(?=-foot)") |> as.numeric(),
    coord_distance = sqrt(coordinate_x^2 + coordinate_y^2),
    shot_distance = case_when(
      !is.na(parsed_distance) ~ parsed_distance,
      str_detect(shot_text, "dunk") ~ 1,
      str_detect(shot_text, "layup|tip shot|tip-in") ~ 2,
      str_detect(shot_text, "hook shot|floating") ~ 6,
      TRUE ~ coord_distance
    )
  ) |>
  filter(
    shooting_play == TRUE,
    !str_detect(shot_text, "free throw"),
    !is.na(athlete_id_1),
    !is.na(team_id),
    !is.na(home_team_id),
    !is.na(away_team_id),
    !is.na(coordinate_x),
    !is.na(coordinate_y),
    !is.na(period_number),
    !is.na(clock_minutes),
    !is.na(clock_seconds),
    !is.na(home_score),
    !is.na(away_score),
    !is.na(shot_distance)
  ) |>
  transmute(
    made = as.integer(scoring_play == TRUE),
    player_id = athlete_id_1,
    team_id = team_id,
    period = period_number,
    is_home = case_when(
      team_id == home_team_id ~ "home",
      team_id == away_team_id ~ "away",
      TRUE ~ NA_character_
    ),
    shot_distance = shot_distance,
    coordinate_x = coordinate_x,
    coordinate_y = coordinate_y,
    shot_type_simple = case_when(
      score_value == 3 ~ "3PT",
      score_value == 2 ~ "2PT",
      str_detect(shot_text, "three point|3-pt|3pt") ~ "3PT",
      TRUE ~ "2PT"
    ),
    shot_zone_basic = case_when(
      shot_distance <= 4 ~ "Restricted Area",
      shot_distance <= 14 ~ "Paint / Short Midrange",
      shot_distance <= 22 ~ "Long Midrange",
      TRUE ~ "Three Point"
    ),
    score_margin = case_when(
      team_id == home_team_id ~ home_score - away_score,
      team_id == away_team_id ~ away_score - home_score,
      TRUE ~ NA_real_
    ),
    clock_seconds_remaining = clock_minutes * 60 + clock_seconds,
    game_date = game_date
  ) |>
  filter(!is.na(is_home), !is.na(score_margin)) |>
  left_join(player_lookup, by = c("player_id" = "athlete_id")) |>
  mutate(
    player_name = coalesce(athlete_display_name, paste("Player", player_id))
  ) |>
  select(
    made,
    player_id,
    player_name,
    period,
    is_home,
    shot_distance,
    coordinate_x,
    coordinate_y,
    shot_type_simple,
    shot_zone_basic,
    score_margin,
    clock_seconds_remaining,
    game_date
  )
```

```{r}
glimpse(shots_raw)
```

# Filtering

To target our primary interest and keep statistical interpretability intact, we next filter down the data to players with at least a "moderate" number of shot attempts and then sample thousands of shots.  This leaves us data for a meaningful model but avoids keeping a huge data set where the posterior intervals are going to be too tight.

```{r}
player_counts <- shots_raw |>
  count(player_name, name = "attempts") |>
  arrange(desc(attempts))

players_keep <- player_counts |>
  filter(attempts >= 60) |>
  slice_head(n = 50) |>
  pull(player_name)

shots_filtered <- shots_raw |>
  filter(player_name %in% players_keep)

shots_model <- shots_filtered |>
  slice_sample(n = min(4000, nrow(shots_filtered))) |>
  mutate(
    distance_z = as.numeric(scale(shot_distance)),
    score_margin_z = as.numeric(scale(score_margin)),
    clock_seconds_z = as.numeric(scale(clock_seconds_remaining)),
    player_name = factor(player_name),
    shot_type_simple = factor(shot_type_simple, levels = c("2PT", "3PT")),
    shot_zone_basic = factor(
      shot_zone_basic,
      levels = c(
        "Restricted Area",
        "Paint / Short Midrange",
        "Long Midrange",
        "Three Point"
      )
    ),
    is_home = factor(is_home, levels = c("away", "home")),
    period = factor(period)
  ) |>
  droplevels()

glimpse(shots_model)
```

# Data Analysis

## Overall make percentage

```{r}
shots_model |>
  summarize(
    n_shots = n(),
    made_shots = sum(made),
    fg_pct = mean(made)
  ) |>
  gt()
```

## Make percentage by shot type

```{r}
shots_model |>
  group_by(shot_type_simple) |>
  summarize(
    n = n(),
    fg_pct = mean(made),
    .groups = "drop"
  ) |>
  gt()
```

## Make percentage by shot zone

```{r}
shots_model |>
  group_by(shot_zone_basic) |>
  summarize(
    n = n(),
    fg_pct = mean(made),
    .groups = "drop"
  ) |>
  arrange(desc(fg_pct)) |>
  gt()
```

## Shot distance and outcome

```{r}
ggplot(shots_model, aes(x = shot_distance, y = made)) +
  geom_jitter(height = 0.05, alpha = 0.08) +
  geom_smooth(
    method = "glm",
    method.args = list(family = binomial),
    se = TRUE
  ) +
  labs(
    title = "Shot outcome by derived shot distance",
    x = "Shot distance",
    y = "Made shot indicator"
  )
```

## Raw field goal percentage by player

```{r}
player_raw <- shots_model |>
  group_by(player_name) |>
  summarize(
    attempts = n(),
    makes = sum(made),
    raw_fg_pct = mean(made),
    .groups = "drop"
  ) |>
  arrange(desc(raw_fg_pct))

player_raw |>
  slice_head(n = 15) |>
  gt()
```

# Bayesian Model

## Likelihood

For shot (i),

$$Y_i \sim \text{Bernoulli}(p_i).$$

## Linear predictor

Our main model is

$$\text{logit}(p_i)
=
\alpha
+
u_{\text{player}[i]}
+
\beta_1 \,\text{distance}_{z,i}
+
\beta_2 I(\text{3PT}_i)
+
\beta_3 I(\text{home}_i)
+
\beta_4 \,\text{period}_i
+
\beta_5 \,\text{clock}_{z,i}
+
\beta_6 \,\text{margin}_{z,i}.$$

The player random effect is

$$u_j \sim \text{Normal}(0, \sigma_{\text{player}}).$$

We interpret (u_j) as the player adjusted shooting ability after controlling for shot distance, type, period, time remaining, score margin, and home/away.

## Priors

We chose uninformative priors.

Our intercept gets:

$$\alpha \sim \text{Normal}(0, 1.5).$$

This prior allows for a wide range of shooting probability priors and since NBA field goal percentages are often 40%-50%, our prior is uninformative but reasonable.

Regression coefficients get:

$$\beta_k \sim \text{Normal}(0, 1).$$

This prior allows for substantial effects while discouraging implausibly extreme effects.  

The player standard deviation gets

$$\sigma_{\text{player}} \sim \text{Exponential}(1).$$

Our prior favors small player-to-player variation but would still allow heterogeneity if the data suggests it.

```{r}
model_formula <- bf(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z + (1 | player_name)
)

model_formula_fixed <- bf(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z
)

priors_main <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sd")
)

priors_fixed_only <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1), class = "b")
)

priors_main
```

# Prior predictive checking

We will now check whether our priors are reasonable in implying shot-making rates.

```{r}
fit_prior <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_main,
  sample_prior = "only",
  chains = 4,
  iter = 1000,
  seed = 415,
  control = list(adapt_delta = 0.95)
)
```

```{r}
pp_check(fit_prior, ndraws = 100) +
  labs(title = "Prior predictive check")
```

# Fitting our Bayesian hierarchical logistic model

```{r}
fit_bayes <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_main,
  sample_prior = "yes",
  chains = 4,
  iter = 2500,
  warmup = 1000,
  seed = 415,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
summary(fit_bayes)
```

# MCMC diagnostics

Now, we are going to check MCMC diagnostic plots.

```{r}
plot(fit_bayes)
```

```{r}
draws_array <- as.array(fit_bayes)

mcmc_trace(
  draws_array,
  pars = c(
    "b_Intercept",
    "b_distance_z",
    "b_clock_seconds_z",
    "b_score_margin_z"
  )
)
```

```{r}
draws_df <- as_draws_df(fit_bayes)

diag_summary <- posterior::summarise_draws(
  draws_df,
  posterior::rhat,
  posterior::ess_bulk,
  posterior::ess_tail
) |>
  rename(parameter = variable)

rhat_col <- names(diag_summary)[str_detect(names(diag_summary), regex("rhat", ignore_case = TRUE))][1]
ess_bulk_col <- names(diag_summary)[str_detect(names(diag_summary), regex("ess_bulk", ignore_case = TRUE))][1]
ess_tail_col <- names(diag_summary)[str_detect(names(diag_summary), regex("ess_tail", ignore_case = TRUE))][1]

diag_summary <- diag_summary |>
  rename(
    rhat_value = all_of(rhat_col),
    ess_bulk_value = all_of(ess_bulk_col),
    ess_tail_value = all_of(ess_tail_col)
  ) |>
  filter(str_detect(parameter, "^(b_|sd_|r_)")) |>
  arrange(desc(rhat_value))

diag_summary |>
  slice_head(n = 20) |>
  gt()
```

A well-fitting model have (\hat{R}) value close to 1.00.

# Posterior inference for fixed effects

```{r}
fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  gt()
```

Our expectation is the distance coefficient should be negative since longer range shots should have lower make probabilities.  The home/away effect, period effect, and game-state effects should be smaller after adjusting for shot distance and player ability.

Interpretation as odds ratios

```{r}
posterior_fixed <- fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("parameter") |>
  mutate(
    odds_ratio_estimate = exp(Estimate),
    odds_ratio_q2.5 = exp(Q2.5),
    odds_ratio_q97.5 = exp(Q97.5)
  )

posterior_fixed |>
  select(parameter, odds_ratio_estimate, odds_ratio_q2.5, odds_ratio_q97.5) |>
  gt()
```

# Player specific posterior estimates

The player effects estimate shooting ability after adjusting for the shot's context.  This is different from raw field goal percent since they are adjusted player effects on the log-odds scale.

```{r}
player_effects <- ranef(fit_bayes)$player_name[, , "Intercept"] |>
  as.data.frame() |>
  rownames_to_column("player_name") |>
  rename(
    player_effect = Estimate,
    lower = Q2.5,
    upper = Q97.5
  ) |>
  arrange(desc(player_effect))

player_effects |>
  slice_head(n = 15) |>
  gt()
```

```{r}
player_effects |>
  slice_max(player_effect, n = 20) |>
  mutate(player_name = fct_reorder(player_name, player_effect)) |>
  ggplot(aes(x = player_effect, y = player_name)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  labs(
    title = "Top adjusted player shooting effects",
    x = "Player random effect on log-odds scale",
    y = NULL
  )
```

# Partial pooling

We will now compare raw field goal percentage with the Bayesian adjusted player effect.  Since players that have fewer shot attempts have higher variability, we expect them to be pulled more strongly towards the league average level.

```{r}
partial_pooling_df <- player_raw |>
  inner_join(player_effects, by = "player_name")

ggplot(partial_pooling_df, aes(x = raw_fg_pct, y = player_effect, size = attempts)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Raw FG% versus Bayesian adjusted player effect",
    subtitle = "Point size represents number of attempts",
    x = "Raw field goal percentage",
    y = "Adjusted player effect on log-odds scale",
    size = "Attempts"
  )
```

# Posterior prediction

Now, we will use the model to predict shot make probabilities for specific shot contexts

```{r}
example_players <- player_effects |>
  slice_head(n = 3) |>
  pull(player_name)

new_shots <- crossing(
  player_name = factor(example_players, levels = levels(shots_model$player_name)),
  shot_distance = c(4, 12, 25)
) |>
  mutate(
    distance_z = (shot_distance - mean(shots_model$shot_distance)) / sd(shots_model$shot_distance),
    shot_type_simple = factor(
      if_else(shot_distance >= 23, "3PT", "2PT"),
      levels = levels(shots_model$shot_type_simple)
    ),
    is_home = factor("home", levels = levels(shots_model$is_home)),
    period = factor(1, levels = levels(shots_model$period)),
    clock_seconds_z = 0,
    score_margin_z = 0
  )
```

```{r}
preds <- fitted(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  scale = "response",
  summary = TRUE
)

bind_cols(new_shots, as.data.frame(preds)) |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    Estimate,
    Q2.5,
    Q97.5
  ) |>
  gt()
```

These predictions are the posterior summaries for the probability of making a shot accounting for uncertainty in the fixed effects and player effects

## Prediction intervals

The posterior predictive distribution is Binary, thus, single shot prediction intervals will be wide and discrete even when the predicted make probability is estiamted precisely.  

```{r}
set.seed(415)


y_pred_single <- posterior_predict(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  ndraws = 4000
)

single_shot_prediction_intervals <- map_dfr(seq_len(ncol(y_pred_single)), function(j) {
  tibble(
    row_id = j,
    predicted_make_rate = mean(y_pred_single[, j]),
    pred_50_lower = quantile(y_pred_single[, j], 0.25),
    pred_50_upper = quantile(y_pred_single[, j], 0.75),
    pred_80_lower = quantile(y_pred_single[, j], 0.10),
    pred_80_upper = quantile(y_pred_single[, j], 0.90),
    pred_98_lower = quantile(y_pred_single[, j], 0.01),
    pred_98_upper = quantile(y_pred_single[, j], 0.99)
  )
})

new_shots |>
  mutate(row_id = row_number()) |>
  left_join(single_shot_prediction_intervals, by = "row_id") |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    predicted_make_rate,
    pred_50_lower,
    pred_50_upper,
    pred_80_lower,
    pred_80_upper,
    pred_98_lower,
    pred_98_upper
  ) |>
  gt() |>
  tab_header(
    title = "Posterior prediction intervals for one future shot",
    subtitle = "Binary posterior predictive outcomes: 0 = miss, 1 = make"
  ) |>
  cols_label(
    predicted_make_rate = "Predicted Pr(make)",
    pred_50_lower = "50% lower",
    pred_50_upper = "50% upper",
    pred_80_lower = "80% lower",
    pred_80_upper = "80% upper",
    pred_98_lower = "98% lower",
    pred_98_upper = "98% upper"
  ) |>
  fmt_percent(columns = predicted_make_rate, decimals = 1)
```

For a single future shot, our prediction intervals serve to answer the question: What outcomes should be expected for one future attempt in this particular context?  Since the observed outcome is binary, these intervals often include 0 and 1, and while that would usually suggest that the model is uninformative, in our context is reflects that even high-probability shots can be missed and low-probability shots can be made.

Now, to make our prediction intervals more interpretable, we simulated the number of made shots out of 100 simulated attempts taken undert the same context to keep the model's uncertainty but show the prediction as a "shots made out of 100" context.

```{r}
set.seed(415)


epred_draws <- posterior_epred(
  fit_bayes,
  newdata = new_shots,
  re_formula = NULL,
  ndraws = 4000
)

made_per_100_draws <- apply(
  epred_draws,
  2,
  function(p_draws) rbinom(length(p_draws), size = 100, prob = p_draws)
)

made_per_100_intervals <- map_dfr(seq_len(ncol(made_per_100_draws)), function(j) {
  tibble(
    row_id = j,
    predicted_makes_per_100 = mean(made_per_100_draws[, j]),
    pred_50_lower = quantile(made_per_100_draws[, j], 0.25),
    pred_50_upper = quantile(made_per_100_draws[, j], 0.75),
    pred_80_lower = quantile(made_per_100_draws[, j], 0.10),
    pred_80_upper = quantile(made_per_100_draws[, j], 0.90),
    pred_98_lower = quantile(made_per_100_draws[, j], 0.01),
    pred_98_upper = quantile(made_per_100_draws[, j], 0.99)
  )
})

new_shots |>
  mutate(row_id = row_number()) |>
  left_join(made_per_100_intervals, by = "row_id") |>
  select(
    player_name,
    shot_distance,
    shot_type_simple,
    predicted_makes_per_100,
    pred_50_lower,
    pred_50_upper,
    pred_80_lower,
    pred_80_upper,
    pred_98_lower,
    pred_98_upper
  ) |>
  gt() |>
  tab_header(
    title = "Posterior prediction intervals for makes out of 100 similar shots",
    subtitle = "50%, 80%, and 98% prediction intervals"
  ) |>
  cols_label(
    predicted_makes_per_100 = "Predicted makes per 100",
    pred_50_lower = "50% lower",
    pred_50_upper = "50% upper",
    pred_80_lower = "80% lower",
    pred_80_upper = "80% upper",
    pred_98_lower = "98% lower",
    pred_98_upper = "98% upper"
  ) |>
  fmt_number(
    columns = c(
      predicted_makes_per_100,
      pred_50_lower,
      pred_50_upper,
      pred_80_lower,
      pred_80_upper,
      pred_98_lower,
      pred_98_upper
    ),
    decimals = 1
  )
```

These intervals that we derived tie back to our original research question.  Our model estimates how shot context and player-specific ability affect make probability and our prediction intervals translate those estimated probabilities into predictions on future results under similar contexts.  Shorter shots should have higher make probabilities than longer shots,  and player with larger positive random effects should have higher prediction intervals than comparable players taking the same shots

# Posterior predictive checking

Now we will use posterior predictive checking to compare simulated data from the model to the observed data.

```{r}
pp_check(fit_bayes, ndraws = 100) +
  labs(title = "Posterior predictive check: overall distribution")
```

Since we have a binary response, we are going to check if the model recreates the group-level summaries like make-percentage by shot zone.

```{r}
yrep <- posterior_predict(fit_bayes, ndraws = 200)

obs_by_zone <- shots_model |>
  group_by(shot_zone_basic) |>
  summarize(obs_fg = mean(made), n = n(), .groups = "drop")

sim_by_zone <- map_dfr(seq_len(nrow(yrep)), function(s) {
  shots_model |>
    mutate(y_sim = yrep[s, ]) |>
    group_by(shot_zone_basic) |>
    summarize(sim_fg = mean(y_sim), .groups = "drop") |>
    mutate(draw = s)
})

zone_ppc <- sim_by_zone |>
  group_by(shot_zone_basic) |>
  summarize(
    sim_lower = quantile(sim_fg, 0.025),
    sim_median = median(sim_fg),
    sim_upper = quantile(sim_fg, 0.975),
    .groups = "drop"
  ) |>
  left_join(obs_by_zone, by = "shot_zone_basic")

ggplot(zone_ppc, aes(x = sim_median, y = fct_reorder(shot_zone_basic, sim_median))) +
  geom_point() +
  geom_errorbarh(aes(xmin = sim_lower, xmax = sim_upper), height = 0.2) +
  geom_point(aes(x = obs_fg), shape = 4, size = 3) +
  labs(
    title = "Posterior predictive check by shot zone",
    subtitle = "Dots are simulated medians; x-marks are observed field goal percentages",
    x = "Field goal percentage",
    y = "Shot zone"
  )
```

We don't see large discrepancies in the model so we accept our model under this context.

# Model comparisons

We compare the hierarchical model to a non-hierarchical Bayesian logistic regression.

```{r}
fit_no_player <- brm(
  formula = model_formula_fixed,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_fixed_only,
  sample_prior = "yes",
  chains = 4,
  iter = 2500,
  warmup = 1000,
  seed = 415,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
loo_bayes <- loo(fit_bayes)
loo_no_player <- loo(fit_no_player)

loo_compare(loo_bayes, loo_no_player)
```

We favor the fit_bayes model since it has a 2.3 advantage over fit_no_player.

# Prior sensitivity analysis

Strong Bayesian analysis should show to what degree conclusions degree on the prior, so we compared our uninformative prior to a tigher prior and a wider prior.

```{r}
priors_tighter <- c(
  prior(normal(0, 1), class = "Intercept"),
  prior(normal(0, 0.5), class = "b"),
  prior(exponential(2), class = "sd")
)

priors_wider <- c(
  prior(normal(0, 3), class = "Intercept"),
  prior(normal(0, 2), class = "b"),
  prior(exponential(0.5), class = "sd")
)
```

```{r}
fit_tighter <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_tighter,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 416,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)

fit_wider <- brm(
  formula = model_formula,
  data = shots_model,
  family = bernoulli(link = "logit"),
  prior = priors_wider,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  seed = 417,
  control = list(adapt_delta = 0.97, max_treedepth = 12)
)
```

```{r}
sensitivity_fixed <- bind_rows(
  fixef(fit_bayes) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "main"),
  fixef(fit_tighter) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "tighter"),
  fixef(fit_wider) |> as.data.frame() |> rownames_to_column("parameter") |> mutate(prior = "wider")
)

sensitivity_fixed |>
  select(prior, parameter, Estimate, Q2.5, Q97.5) |>
  gt()
```

# Frequentist methods comparison

Now, we will fit a logistic mixed-effects model and compare to our Bayesian analysis.

```{r}
fit_freq <- glmer(
  made ~ distance_z + shot_type_simple + is_home + period +
    clock_seconds_z + score_margin_z + (1 | player_name),
  data = shots_model,
  family = binomial(link = "logit"),
  control = glmerControl(optimizer = "bobyqa")
)
```

```{r}
summary(fit_freq)
```

```{r}
freq_fixed <- tidy(fit_freq, effects = "fixed", conf.int = TRUE)

bayes_fixed <- fixef(fit_bayes) |>
  as.data.frame() |>
  rownames_to_column("term") |>
  mutate(term = if_else(term == "Intercept", "(Intercept)", term)) |>
  rename(
    bayes_estimate = Estimate,
    bayes_lower = Q2.5,
    bayes_upper = Q97.5
  )

comparison <- freq_fixed |>
  select(term, freq_estimate = estimate, freq_lower = conf.low, freq_upper = conf.high) |>
  left_join(bayes_fixed, by = "term")

comparison |>
  gt()
```

The frequentist logistic model estimates similar coefficients, but our Bayesian model provides complete posterior distributions that allow for more interpretable probability statements.  Also, the Bayesian method creates more natural workflows to perform prior predictive checks and prior sensitivity analysis.

# Results

Initially, our data exploration yielded the findings that shot distance is strongly correlated with shooting success.  Restricted area shots had estimated field goal percentage of 66% compared to 49% for threes.  Similarly, two pointers had a field goal percentage of 50% while threes had a field goal percentage of 42% indicating that closer shots are much more likely to be made than long-range shots.  To investigate this relationship further, we fir a hierarchical Bayesian logistic regression model that concluded that the most important predictors of shot success are shot distance and score margin.  

Shot distance was the strongest negative predictor with odds ratio 0.752 (95% credible interval: 0.698, 0.811) meaning that a one standard deviation increase in distance reduced the odds of making a shot by ~25%.

Score margin was the strongest positive predictor.  The odds ratio was 1.325 (95% credible interval: 1.239, 1.419) indicating that larger scoring margins increased the odds of making a shot by ~33%.

Other variables like shot type, home/away, clock time, and game periods did not end up showing strong evidence of meaningful effects after accounting for distance and scoring margin.

We also used our model to estimate player-specific effects and did so by taking players like Shai Gilgeous-Alexander, Pascal Siakam, Mikal Bridges Donovan Mitchell, among others with positive estimated shooting effects and calculating credible intervals for their make percentage.  However, all of the players' credible intervals contained 0 which suggests that we have low certainty that, based on model results, we can distinguish these players from league-average shooters.

Posterior prediction tables showed that players were expected to make substantially more close-range two-point shots than long-range three-point shots. For example, predicted makes out of 100 attempts consistently decreased as shot distance increased. Predictions for individual shots remained highly uncertain because each shot can only result in a make or miss.

Then, we conducted a model comparison that found that our hierarchical model was more effective than a non-hierarchical alternative which suggests that player-specific effects are at least moderately helpful in improving model predictive performance.

# Conclusion

The hierarchical Bayesian logistic regression model that we fit is effective at explaining NBA shot-making while accoiunting for variables like player differences and shot context.  We found through this investigation that greater shot distance substantially lowers shot-making probabilities while higher score margins increase it.  We also found that there are player-to-player variations, however, our model could not confidently identify them, but overall our model provides a strong representation of some of the major factors that influence NBA shot success.
